Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway

### 2021-08-03 analysis/paper/paper.Rmd
Minor edits to references and readme

---
title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Department of Archaeology, History and Conservation
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely. The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods. Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g. Clark and Barton 2017), originally devised for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting. Furthermore, this finding supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.

@@

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [@fuglestvedt2007; @eigeland2015; @mansrud2016]. These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

@@

The exploratory approach taken here means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported upon, the data and employed R programming script [@rcoreteam] is freely available as a research compendium at https://osf.io/ehjfc/, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].

@@

```{r var-contrib, fig.cap = "Contribution of variables to the components of the PCA. The dotted red line indicates the expected contribution from each variable given a uniform distribution of impact.", fig.height = 3}
# The plots below investigates the contribution of the variables to the
# components. The code has in large part been taken from 
# Bicho & Cascalheira (2020).
var <- get_pca_var(pcalithics)
dim1 <- fviz_contrib(pcalithics, choice = "var", axes = 1,
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 1")
dim2 <- fviz_contrib(pcalithics, choice = "var", axes = 2, 
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 2")
dim3 <- fviz_contrib(pcalithics, choice = "var", axes = 3,
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 3")
dim4 <- fviz_contrib(pcalithics, choice = "var", axes = 4,
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 4")

@@

The results of the CA appear to align well with previous research [e.g. @solheim2017c, with references]. In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [cf. @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b]. The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [@damlien2016]. When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones. The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of non-flint hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016]. However, in agreement with the literature, this is evidently not as prominent a part of these assemblages.

@@

<!-- ```{r colophon, cache = FALSE} -->
<!-- # which R packages and versions? -->
<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
<!-- ``` -->

@@

<!-- ```{r} -->
<!-- # what commit is this file at?  -->
<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
<!-- ``` -->

>>>>

---
title: "Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Department of Archaeology, History and Conservation
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely. The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods. Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g. Clark and Barton 2017), originally devised for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting. Furthermore, this finding supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.

@@

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g. @breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g. @fuglestvedt2007; @eigeland2015; @mansrud2016]. These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

@@

The exploratory approach taken here means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported upon, the data and employed R programming script [@rcoreteam] is freely available as a research compendium at <https://osf.io/ehjfc/>, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].

@@

```{r var-contrib, fig.cap = "Contribution of variables to the first two components of the PCA. The dotted red line indicates the expected contribution from each variable given a uniform distribution of impact.", fig.height = 3}
# The plots below investigates the contribution of the variables to the
# components. The code has in large part been taken from 
# Bicho & Cascalheira (2020).
var <- get_pca_var(pcalithics)
dim1 <- fviz_contrib(pcalithics, choice = "var", axes = 1,
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 1")
dim2 <- fviz_contrib(pcalithics, choice = "var", axes = 2, 
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 2")
dim3 <- fviz_contrib(pcalithics, choice = "var", axes = 3,
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 3")
dim4 <- fviz_contrib(pcalithics, choice = "var", axes = 4,
                     color = "black", fill = "gold2", 
                     title = "Contribution of variables to Component 4")

@@

The results of the CA appear to align well with previous research [e.g. @solheim2017c, with references]. In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [cf. @bjerck2017; @breivik2018; @fuglestvedt2009; @jaksland2014b; @damlien2018b]. The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [@damlien2016]. When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones. The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of non-flint hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016]. However, in agreement with the literature, this is evidently not as prominent a part of these assemblages.

@@

<!-- ```{r colophon, cache = FALSE} -->

<!-- # which R packages and versions? -->

<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->

<!-- ``` -->

@@

<!-- ```{r} -->

<!-- # what commit is this file at?  -->

<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->

<!-- ``` -->
